Gravitational field calculation in spherical coordinates using variable densities in depth

### 2018-09-21 manuscript/manuscript.tex
Finish the linear density analysis section

\title[Variable Density Tesseroids]{
    Gravitational field calculation in spherical coordinates using variable
    densities in depth
}
\author[S.R. Soler, A. Pesce, L. Uieda and M.E. Gimenez]{
    Santiado R. Soler$^{1,2}$, Agustina Pesce$^{1,2}$, Leonardo Uieda$^3$ and
    Mario E. Gimenez$^{1,2}$ \\
    $^1$CONICET, Argentina. e-mail: santiago.r.soler@gmail.com\\
    $^2$Instituto Geofísico Sismológico Volponi, Universidad Nacional de
    San Juan, Argentina\\
    $^3$Department of Geology and Geophysics, SOEST, University of Hawai‘i at
    M\={a}noa, Honolulu, Hawaii, USA
}

@@

@@

\begin{equation}
    \rho(r') = ar' + b,
\end{equation}

@@

The absolute density difference defined on equation
\ref{eq:density-abs-diff} will always be zero for the linear density case.
As a result, the inequality \ref{eq:delta-density} will always be satisfied and no
divisions will ever be performed.
Therefore, the distance-size ration $D$ of the adaptive discretization algorithm is the
only mechanism that controls the accuracy of the numerical integration.
For this reason, we will only determine the minimum value of $D$ needed in order to
guarantee an acceptable accuracy while ignoring the value of $\delta$.

In order to compare the numerical model with the analytical solution we
must build a spherical shell made of tesseroids.
We use a mesh of $30^\circ \times 30^\circ$ tesseroids whose outer and
inner radii are the same ones as the spherical shell in order to compute
the gravitational potential, the vertical component of the gradient
($g_z$), and the diagonal components of the Marussi tensor ($g_{xx}$,
$g_{yy}$, $g_{zz}$) on four different grids: (1) a grid located at the
pole, (2) another one on the Equator, (3) one located at satellite
height and (4) a big grid of $30^\circ \times 30^\circ$.
More information about this grids can be found on Table
\ref{tab:grids}.

We will compute these gravitational fields on every point of the four
grids for different values of $D$, ranging from 0.5 to 10 with a step
of 0.5.
Then we will calculate the maximum difference between these results and the
value that the analytical solution assumes at the same height of the
computation grid.
Because the gravity fields generated by a tesseroid are approximated by
the effect of point masses, the numerical results may vary between
computation points at same height but at different latitude or
longitude.
That's why we will choose for comparison the maximum difference between the
results on every point of the grid and the analytical solution.

Finally, we will set the acceptable value of $D$ as the minimum value
at which the corresponding error of the numerical approximation is
lower than 0.1\%.

\begin{table}
\caption{
    Description of the synthetic grids on which the comparison of the
    numerical model against the analytical solutions for a spherical
    shell was done.
    This collection of grids tries to include every possible scenario
    on which the numerical model can present a different behaviour
    depending on the size of the grid, its location or its height over
    the Earth surface. Every grid consists in a set of 10$\times$10
    points.
}
\label{tab:grids}
\begin{tabular}{lcccc}
    Grid & Size & Lat. extension & Lon. extension & Height \\ \hline
    Pole & $1^\circ \times 1^\circ$ & $89^\circ - 90^\circ$ &
        $0^\circ - 1^\circ$ & 2km \\
    Equator & $1^\circ \times 1^\circ$ & $0^\circ - 1^\circ$ &
        $0^\circ - 1^\circ$ & 2km \\
    Satellite & $1^\circ \times 1^\circ$ & $89^\circ - 90^\circ$ &
        $0^\circ - 1^\circ$ & 260km \\
    Big Grid & $30^\circ \times 30^\circ$ & $60^\circ - 90^\circ$ &
        $0^\circ - 30^\circ$ & 2km \\
\end{tabular}
\end{table}

In order to perform the comparison, we proposed a thin and a thick
spherical shells with 1km and 35km of thickness, respectively, whose
outer radii are equal to the mean Earth radius.
On both cases we will suppose that the density assumes a value of
2670~kg/m$^3$ on the outer surface and 3300~kg/m$^3$ at the inner
radius:

\begin{equation}
    \rho(r') = ar' + c,
    \label{eq:density-linear}
\end{equation}
\noindent with
\begin{equation}
    a = -\frac{3300\text{kg/m$^3$} - 2670\text{kg/m$^3$}}{R_2 - R_1},
\end{equation}
\begin{equation}
    c = \frac{3300\text{kg/m$^3$} -
        2670\text{kg/m$^3$}}{R_2 - R_1} R +
        2670\text{kg/m$^3$},
\end{equation}

\noindent where $R$ = 6378.137 km is the mean Earth radius.

We have computed the percentage difference between the numerical model
with its corresponding analytical solution (equation
\ref{eq:shell-pot-linear}) for the thick and the thin spherical shell
on each of the four grids described in Table \ref{tab:grids}.
We can see the results for the potential, the vertical component of the
gradient $g_z$ and the $g_{zz}$ component of the Marussi tensor on
Figure \ref{fig:D-linear}.
We exclude the results corresponding to the other two diagonal
components of the tensor because they are similar to the ones obtained
for the $g_{zz}$, although they can be found on the repository.
\todo[inline]{chequear}

For the linear density case, we can observe that the calculation of the
potential and the $g_z$ component of its gradient needs a value of
$D=1$ and $D=2$, respectively, in order to ensure a numerical error
bellow the 0.1\% threshold.
On the other hand, the $g_{zz}$ computations show a noticeable
difference between the thin and the thick shell: the first one needs a
value of $D$ equal to 8 while the later only a $D$ of 3.5.
We will keep the more conservative result of $D$ equal to 8 in case of
computing the Marussi tensor components.

\begin{figure*}
\centering
\includegraphics[width=\linewidth]{figures/linear-D.pdf}
\caption{
    Differences between the gravity fields generated by the numerical model
    and the analytical solution for a thin and a thick spherical shell of 1km
    and 35km of thickness, respectively.
    Both models have the linear density defined on equation
    \ref{eq:density-linear}.
    The computations were performed on the four grids described in
    Table \ref{tab:grids}, for different values of the distance-size
    ratio $D$.
    Due to the linearity of the density function, the density-based
    discretization algorithm is not applied.
    If the difference is less than 0.1\%, we consider that the model
    has achieved an acceptable accuracy for the corresponding value of
    $D$.
    }
\label{fig:D-linear}
\end{figure*}

>>>>

\title[Variable Density Tesseroids]{
    Gravitational field calculation in spherical coordinates using variable
    densities in depth
}
\author[S.R. Soler, A. Pesce, L. Uieda and M.E. Gimenez]{
    Santiado R. Soler$^{1,2}$, Agustina Pesce$^{1,2}$, Leonardo Uieda$^3$ and
    Mario E. Gimenez$^{1,2}$ \\
    $^1$CONICET, Argentina. e-mail: santiago.r.soler@gmail.com\\
    $^2$Instituto Geofísico Sismológico Volponi, Universidad Nacional de
    San Juan, Argentina\\
    $^3$Department of Earth Sciences, SOEST, University of Hawai‘i at
    M\={a}noa, Honolulu, Hawaii, USA
}

@@

\begin{table}
\caption{
    Description of the tesseroid models of a spherical shell and computation grids
    used to characterize the accuracy of the numerical integration.
    Every grids consist of a set of $10 \times 10$ points.
}
\label{tab:grids}
\begin{tabular}{lccc}
    Name & Tesseroid size & Grid region (degrees) & Grid height (km)
    \\ \hline
    Pole & $1^\circ \times 1^\circ$ & $0\text{E}/1\text{E}/89\text{N}/90\text{N}$ & 2 \\
    Equator & $1^\circ \times 1^\circ$ & $0\text{E}/1\text{E}/0\text{N}/1\text{N}$ & 2 \\
    Satellite & $1^\circ \times 1^\circ$ & $0\text{E}/1\text{E}/89\text{N}/90\text{N}$ & 260 \\
    Big Grid & $30^\circ \times 30^\circ$ & $0\text{E}/30\text{E}/60\text{N}/90\text{N}$ & 2 \\
\end{tabular}
\end{table}

@@

\begin{equation}
    \rho(r') = ar' + b,
    \label{eq:density-linear}
\end{equation}

@@

The absolute density difference defined on equation
\ref{eq:density-abs-diff} will always be zero for the linear density case.
As a result, the inequality \ref{eq:delta-density} will always be satisfied and no
divisions will ever be performed.
Therefore, the distance-size ratio $D$ of the adaptive discretization algorithm is the
only mechanism that controls the accuracy of the numerical integration.
For this reason, we will only determine the minimum value of $D$ needed in order to
guarantee an acceptable accuracy while ignoring the value of $\delta$.

In order to compare the numerical results with the analytical solution we
must build a spherical shell made of tesseroids.
We use a combination of tesseroid models with different resolutions and compute their
effects on $10 \times 10$ point grids at different heights and locations.
All combinations of models and grids are shown in Table \ref{tab:grids}.
The tesseroid model is composed of a single radial layer.
The combinations in Table \ref{tab:grids} are repeated for two shells with thickness 1
km and 35 km.
The outer radius of both shells are equal to the mean Earth radius and
the density varies linearly (Eq. \ref{eq:density-linear})
with angular coefficient

\begin{equation}
    a = -\frac{3300\text{kg/m$^3$} - 2670\text{kg/m$^3$}}{R - R_1},
\end{equation}

\noindent and linear coefficient

\begin{equation}
    c = \frac{3300\text{kg/m$^3$} -
        2670\text{kg/m$^3$}}{R - R_1} R +
        2670\text{kg/m$^3$},
\end{equation}

\noindent
in which $R$ = 6378.137 km is the mean Earth radius and $R_1$ is the inner radius of the
shell (determined by the thickness).

We compute the gravitational potential ($V$), the vertical component of the gradient
($g_z$), and the diagonal components of the Marussi tensor ($g_{xx}$, $g_{yy}$,
$g_{zz}$) for each of the four tesseroid models and grids in Table
\ref{tab:grids}.
Other components are equal to zero outside of the shell and are thus omitted from the
analysis.
The computations are repeated for values of $D$ ranging from 0.5 to 10 with a step of
0.5.
We then will calculate the maximum absolute difference between these results and the
analytical solutions derived in Appendix \ref{sec:shell}.
The differences are shown in Fig. \ref{fig:D-linear} as relative to the spherical shell
values.
We omit the differences for $g_{xx}$ and $g_{yy}$ because they are equivalent to the
ones obtained for $g_{zz}$.
Finally, we set the optimal value of $D$ as the minimum value at which the corresponding
error of the numerical approximation is lower than 0.1\%.

We observe from Fig. \ref{fig:D-linear} that the relative errors for the
potential and the $g_z$ component fall below the 0.1\% threshold at
$D=1$ and $D=2$, respectively, for both shell thicknesses.
On the other hand, $g_{zz}$ displays a noticeable
difference between the thin and the thick shell: the first one needs a
value of $D$ equal to 8 while the later only a $D$ of 3.

\begin{figure*}
\centering
\includegraphics[width=\linewidth]{figures/linear-D.pdf}
\caption{
    Differences between the gravitational fields generated by the tesseroid model
    and the analytical solution for a thin (left) and a thick (right) spherical shell.
    Both models have a linear density function of radius (Eq. \ref{eq:density-linear}).
    The computations were performed on the four combinations described in
    Table \ref{tab:grids}.
    Due to the linearity of the density function, the density-based discretization
    algorithm is not applied.
    Differences are reported as a percentage of the shell values.
    The horizontal dashed black line represents a target difference of 0.1\%.
}
\label{fig:D-linear}
\end{figure*}
